Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-03-21 paper/paper.tex
Fixed typos and restructured figures

	The structure of this paper is as follows.
	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
	In section~\ref{discussion} we propose a new parallel spatial locality metric, implement it in the AIWC framework, and evaluate it over the matrix multiplication example from section~\ref{motivating example}.
	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial metric for OpenCL codes.
	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.

@@

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after this optimisation.

@@

	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimisations to the simple matrix multiply kernel.

@@

	\flushbottom
	\section{Related Work} \label{related work}

@@

	\begin{figure*}[ht!]
		\centering
		\subfloat{
			\centering
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT},%alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time (ms)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
			legend style={at={(0.6,0.45)},anchor=west}
			]

@@

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric obtained from AIWC for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Normalised parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
		\label{fig: opendwarfs parallel spatial locality}
	\end{figure*}

@@

	At each point in the kernels' memory access profile, each work-item with global ID $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices. The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching \cite{krommydas2016opendwarfs}. However, memory requests made simultaneously by a work-group always fall within a rectangular block of one of the matrices. 
	This allows memory access coalescing on GPUs for OpenCL work-items accessing contiguous matrix data, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. 
	However, work-items accessing data along a column of the matrix do not observe memory access coalescing.
	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.

	\subsection{Graph Traversal: BFS}

@@

	Of the two BFS kernels, it is \texttt{BFS kernel 1} that performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.

	\subsection{Sparse Linear Algebra: CSR}

@@

	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work item. 
	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing.
	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed. The collected parallel spatial locality metric reflects this trend. Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.

	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}

	 The recorded parallel spatial locality in GEM starts at a low value of 0.08604 and approaches the theoretical lower limit of 0 (2.18e-4 at 10 bits dropped). This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads. From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronised threads on the GPU.

	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.

	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wuntsch, we observe a flat parallel spatial locality graph. 
	 The worst possible memory access pattern: all threads in a workgroup accessing results in a flat PSL metric of 1. 
	 The PSL metric for Needleman-Wuntsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
	 However, the workload imbalance across threads in each OpenCL work group eases pressure on the memory unit when not all threads are active.

	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow dropoff in the PSL metric at low $n$ values. This is reflective of the mostly disparate parallel memory access pattern in these codes. The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localisation of memory accesses.

@@

	- 
	\end{comment}

	\section{Conclusions and Future Work} \label{future work}

	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to characterise hardware-specific optimization. 
	We implemented and evaluated a new feature, relative local memory usage, to help characterize memory-based GPU code optimizations.
	We also implemented a new parallel spatial locality metric to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
	We ran the enhanced AIWC tool on matrix multiply kernels and selected OpenDwarfs benchmarks, presenting results and analysis to validate our methodology. 
	%Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner. %Josh, Beau: is this too bold a claim? If unnecessary, please remove.
	%TODO UPDATE IN FUTURE

>>>>

	The structure of this paper is as follows.
	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion and also consider relevant architecture-dependent approaches to memory access profiling.
	In section \ref{method} we consider non-parallel memory metrics collected by AIWC and evaluate the suitability of these metrics for capturing the impact of performance optimizations to various OpenCL codes.
	In section~\ref{discussion} we propose a new parallel spatial locality metric, implement it in the AIWC framework, and evaluate it over the matrix multiplication example from section~\ref{motivating example}.
	In section~\ref{results} we present the collected metric for selected benchmarks from the Extended OpenDwarfs benchmark suite to validate our methodology in this paper.
	In section~\ref{guidelines} we summarize our findings and present guidelines for interpreting the new parallel spatial locality metric for OpenCL codes.
	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.

@@

	\begin{figure*}[h!]
		\centering
		\subfloat{
			\centering
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT},%alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time (ms)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=80\ \ \ \ $,$N=256\ \ $},
			legend style={at={(0.6,0.45)},anchor=west}
			]

@@

	The code is further optimized by improving the locality of reads from Matrix $B$.
	While in the previous kernel each thread reads only a single element of matrix $A$ from global memory, each thread reads a full column of matrix $B$.
	The repeated reads of elements of matrix $B$ can be shared between threads in a work group by reading tiles from matrix $B$ into shared memory.
	Appendix~\ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.
	Performance is further improved to 614 GLOP/s after this optimization.

@@

	We achieve 1061 GFLOP/s after applying all the recommended GPU memory optimizations to the simple matrix multiply kernel.

@@

	\section{Related Work} \label{related work}

@@

@@

	\flushbottom

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Normalized parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
		\label{fig: opendwarfs parallel spatial locality}
	\end{figure*}

@@

	At each point in the kernels' memory access profile, each work-item with global ID $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices. The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching \cite{krommydas2016opendwarfs}. However, memory requests made simultaneously by a work-group always fall within a rectangular block of one of the matrices. 
	This allows memory access coalescing on GPUs for OpenCL work-items accessing contiguous matrix data, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. 
	However, work-items accessing data along a column of the matrix do not observe memory access coalescing.
	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.
	\subsection{Graph Traversal: BFS}

@@

	Of the two BFS kernels, it is \texttt{BFS kernel 1} that performs the graph traversal necessary to generate a list of neighbours at each node. The memory access patterns in \texttt{BFS kernel 1} are irregular. 
	Work-items fetch discontinuous memory locations attributed to any particular node in the graph, depending on the connectivity of the node they operate on. 
	The program structure involves multiple levels of pointer-chasing and thus the precise memory addresses accessed are data-dependent. 
	This leads to poor parallel spatial locality. Thus, we see a slow dropoff in parallel spatial locality for \texttt{BFS kernel 1}.
	\subsection{Sparse Linear Algebra: CSR}

@@

	In the OpenDwarfs implementation, each row of the input sparse matrix is assigned to a separate OpenCL work item. 
	The locations of matrix data read by each work-item are dependent on the number and position of non-zero values in the sparse-matrix, which are decided by the values in $Aj$ and $Ap$. 
	Similar to BFS, this means memory access patterns are runtime-dependent due to indirect addressing.
	This pattern of indirect addressing is typical of applications in the Spare Linear Algebra Dwarf, which severely hinders locality of memory accesses performed. The collected parallel spatial locality metric reflects this trend. Figure \ref{fig: opendwarfs parallel spatial locality} shows the gradual decline in parallel spatial locality as the number of bits dropped is increased.
	\section{Guidelines: Using the Parallel Spatial Locality Metric} \label{guidelines}

	 The recorded parallel spatial locality in GEM starts at a low value of 0.08604 and approaches the theoretical lower limit of 0 (2.18e-4 at 10 bits dropped). This indicates that almost all memory accesses made by the kernel are perfectly synchronized between OpenCL threads. From a memory performance standpoint, this is the best possible scenario as single memory accesses service all the active synchronized threads on the GPU.

	 Codes with a mix of favourable and unfavourable access patterns (typically involving two-dimensional matrices such as in LUD and SRAD) will have a steep dropoff in PSL to about half of the starting PSL metric as $n$ increases, and the curve flattens out at that point.

	 In the extreme case, when parallel memory accesses are entirely disparate as for Needleman-Wunsch, we observe a flat parallel spatial locality graph. 
	 The worst possible memory access pattern occurs when all threads in a workgroup access disparate memory locations. This results in a flat PSL metric of 1. 
	 The PSL metric for Needleman-Wunsch is a flat line at 0.53. This indicates that parallel memory accesses are entirely disparate. 
	 However, the workload imbalance across threads in each OpenCL work group eases pressure on the memory unit when not all threads are active.

	 Codes with irregular or data-driven memory access patterns like CSR and BFS in the OpenDwarfs Suite have a slow dropoff in the PSL metric at low $n$ values. This is reflective of the mostly disparate parallel memory access pattern in these codes. The increasing steepness of the dropoff at higher $n$ in codes with data-driven memory accesses is reflective of the size of buffers being operated upon as the data structures used during computation cause localization of memory accesses.

@@

	- 
	\end{comment}
	\section{Conclusions and Future Work} \label{future work}

	To the best of our knowledge, this work is the first to propose the use of architecture-independent metrics of parallel memory access to characterize hardware-specific optimization. 
	We implemented and evaluated a new feature, relative local memory usage, to help characterize memory-based GPU code optimizations.
	We also implemented a new parallel spatial locality metric to capture the idea of closeness of memory accesses made by parallel OpenCL workloads.
	We ran the enhanced AIWC tool on matrix multiply kernels and selected OpenDwarfs benchmarks, presenting results and analysis to validate our methodology. 
	%Our work is the first to measure memory access patterns of parallel codes in an architecture-independent, easily interpreted manner. %Josh, Beau: is this too bold a claim? If unnecessary, please remove.
	%TODO UPDATE IN FUTURE
